SPAGHETTI Optimizes the Quantitative Analysis of \textit{In Vitro} Phase Contrast Miscroscopy through a Generative Deep Learning Model

### 2024-11-26 paper.tex
finish up current methods and start intro

@@

\subsection*{LIVECell Cell Type Classification}

\subsection*{C2C12 Cell Media Condition Classification}

>>>>

Phase contrast microscopy (PCM) is an old yet widely used techniques in \textit{in vitro} cell biology research. Comparing to other non-stained methods such as bright-field microscopy, PCM provides a more detailed and contrasted image of the cells by converting the phase shift of the light passing through the cells into brightness changes \autocite{pcm_background}. Its ability to visualize the internal structures of the cells without the need of staining also makes it a popular choice for quick live-cell imaging and temporal cell culture monitoring as long time monitoring of fluorescence stains are known to cause phototoxicity in the cells \autocite{phototoxicity}. However, despite its popularity in cell state monitoring, PCM images are rarely used in analysis routines by biologists. This is because even though they can provide a lot of information about the cells, few attempts have been made to extract this information quantitatively, making biologists relying on assays measurments and readings from stain intensities to investigate their properties of interests. This not only adds additional steps and costs to the experiments, but also limits the amount of information that can be extracted from the images as assays and stain intensities can only measure very limited number of properties at once.

Similar problems have been faced by pathologists in the analysis of histopathology images before. Hematoxylin and eosin (H\&E) staining is the most common staining method used in pathology to visualize the tissue structures and cell morphology and can provide a lot of rich information about the tissue disease state. However, for a long time, the information that can be extracted from H\&E images is limited to the human eye, and the analysis of H\&E images is often qualitative and subjective \autocite{he_stain}. Thanks to the development of deep learning, various approaches have been proposed to develop foundational feature extractors to extract the features from H\&E images and use them for downstream tasks such as making diagnosis and survival prediction. Recently, some attempts have even been made to make inferences of the gene expressions from H\&E images, suggesting a vast amount of information can be extracted by deep learning by looking at the cellular morphologies to make H\&E analysis more quantitative. \todo{add citations}

@@

\subsection*{LIVECell Cell Type Classification}
We performed cell type classification on the 20\% testing LIVECell dataset by training a random forest classifer with 100 trees and a maximum depth of 2. We utilized both Phikon-v2 and H-Optimus-0 \autocite{hoptimus0}, two state-of-the-art foundational models for H\&E slides, as feature extractors, as well as a ResNet model pre-trained on ImageNet as the baseline feature extractor. The task was to utilize the features extracted from the feature extractors to classify the image into one of the eight cell types annotated by the LIVECell dataset. We reported the AUC values from the ROC plots and their standard deviations of the random forest classifiers using a 10-fold cross-validation approach. We repeated the above process for using SPAGHETTI as the pre-processing step, using the original cycleGAN architecture as the pre-processing step, and using the original PCM images before feeding the images into the feature extractors. We performed two-tailed paired t-tests to compare the classification accuracies between the different pre-processing steps. \todo{check if the stats test is necessary}

\subsection*{C2C12 Cell Media Condition Classification}
To validate the performance of SPAGHETTI outside of the training dataset, we performed cell media condition classification on the C2C12 PCM dataset \autocite{c2c12}. The C2C12 dataset consists of PCM images of C2C12 cells under different culture media conditions, which are: control (no growth factors), fibroblast growth factor 2, bone morphogenetic protein 2, and a combination of fibroblast growth factor 2 and bone morphogenetic protein 2. Similar to classifying cell types for LIVECell, we utilized Phikon-v2 and H-Optimus-0 as our H\&E feature extrators and pre-trained ResNet as our baseline feature extractor and trained a random forest classifier with 100 trees and a maximum depth of 2 to classify the images from the extracted features. We reported the AUC values from the ROC plots and their standard deviations of the random forest classifiers using a 10-fold cross-validation approach. We repeated the above process for using SPAGHETTI as the pre-processing step, using the original cycleGAN architecture as the pre-processing step, and using the original PCM images before feeding the images into the random forest classifier. We performed two-tailed paired t-tests to compare the classification accuracies between the different pre-processing steps. \todo{check if the stats test is necessary}
